---
title: "Unsupervised Mini Project"
author: "Sanket Garg (A59026686)"
format: pdf
---

# Preparing the Data
```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names = 1)
```

Removing the First column since it is not required.
```{r}
wisc.data <- wisc.df[,-1]
```

Storing Diagnosis as a separate variable.
```{r}
diagnosis <- as.factor(wisc.df[,1])

diagnosis
```

Q1. How many observations are in this dataset?
```{r}
dim(wisc.data)
```
There are 569 observations in this dataset.

Q2. How many of the observations have a malignant diagnosis?
```{r}
sum(diagnosis == "M")
```
212 observations have a malignant diagnosis.

Q3. How many variables/features in the data are suffixed with _mean?
```{r}
grep("_mean", colnames(wisc.data))
```
The first 10 columns names in the data are suffixed with _mean.


# 2. Principal Component Analysis

Checking column means and standard deviations:
```{r}
colMeans(wisc.data)

apply(wisc.data, 2, sd)
```
Perform PCA on wisc.data
```{r}
wisc.pr <-prcomp(wisc.data, scale. = TRUE)
summary(wisc.pr)
```
Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

0.4427 proportion of the original variance is captured by PC1.

Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Only the first 3 PCs are required to describe at least 70% of the original variance in the data.

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

The first 7 PCs are required to describe at least 90% of the original variance in the data.

Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is difficult to make sense of this plot because it is highly clustered and difficult to visualize individual data points. 

Making a biplot:
```{r}
biplot(wisc.pr)
```
Scatter plot of PC1 vs PC2
```{r}
plot(wisc.pr$x[,1:2], col = diagnosis, xlab="PC1", ylab="PC2")
```
Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

The Malignant ones are relatively well separated from the Benign ones.
Scatter plot for PC1 vs PC3:
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis, xlab="PC1", ylab="PC3")
```
Using ggplot for fancier plots:
```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

library(ggplot2)

ggplot(df) +
  aes(x = df$PC1, y=df$PC2, col=df$diagnosis) +
  geom_point() +
  labs(x = "PC1", y="PC2")
```

Calculate variance of each component:
```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
Variance explained by each component:
```{r}
pve <- pr.var/sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```
Scree plot in the form of bars:
```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
Using the factoextra package:
```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```
Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.
```{r}
wisc.pr$rotation["concave.points_mean",1]
```

#3. Hierarchical Clustering

Scale the data:
```{r}
data.scaled <- scale(wisc.data)
```

Calculate the Euclidean distance between all pairs of observations:
```{r}
data.dist <- dist(data.scaled)
```

Perform the clustering:
```{r}
wisc.hclust <- hclust(data.dist, "complete")
```


Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19.5, col="red", lty=2)
```
Use cutree() to cut the tree so that it has 4 clusters
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

Using table
```{r}
table(wisc.hclust.clusters, diagnosis)
```
Q11. Trying with a different number of clusters
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=8)
table(wisc.hclust.clusters, diagnosis)

```
Q12. Which method gives your favorite results for the same data.dist data set? Explain your reasoning.
```{r}
wisc.hclust_new <- hclust(data.dist, "ward.D2")
wisc.hclust_new.clusters <- cutree(wisc.hclust_new, k=4)
table(wisc.hclust_new.clusters, diagnosis)
```
For grouping into 4 clusters, the "ward.D2"  and the "complete" methods generate the best results for a total of 4 clusters.


#4. Combining the Results

Clustering on PCA results
```{r}
wisc.pr.dist <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(wisc.pr.dist, "ward.D2")
plot(wisc.pr.hclust)
```
Checking the clustering results:
```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
table(grps, diagnosis)
```
Plotting based on the grps
```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
plot(wisc.pr$x[,1:2], col=g)
```
Plotting based on diagnosis variable
```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)

```
Q13. How well does the newly created model with four clusters separate out the two diagnoses?
```{r}
new_grps <- cutree(wisc.pr.hclust, k=4)
table(new_grps, diagnosis)
```
Q14. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.
```{r}
a <- table(wisc.hclust_new.clusters, diagnosis)
a
```


#6. Prediction
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
Q16. Which of these new patients should we prioritize for follow up based on your results?

Based on the results, we should prioritize the patient 2 for follow up based on the results.
