---
title: "Class 08 Mini Project"
author: "A59026686"
format: html
---
# Importing Candy Data
```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy = read.csv(candy_file, row.names=1)
```

Q1. How many different candy types are in this dataset?
```{r}
dim(candy)
```
There are 85 types of candies.

Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity)
candy[as.logical(candy$fruity),]
```
There are 38 fruity type candies.

# 2. What is your favourite candy?
Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Hershey's Kisses",]$winpercent
```
Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat",]$winpercent
```
Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
The skim function:
```{r}
library("skimr")
skim_candy <- skim(candy)
```
Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

All the variables are between 0 and 1, but the "winpercent" variable is between 0 and 100.

Q7. What do you think a zero and one represent for the candy$chocolate column?
It represents which are candies are chocolaty and which all aren't.

Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```
Q9. Is the distribution of winpercent values symmetrical?

No, it is not symmetrical. 

Q10. Is the center of the distribution above or below 50%?
```{r}
median(candy$winpercent)
```
The center of the distribution is below 50%.

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
chocolate_mean <- candy$winpercent[as.logical(candy$chocolate)]
fruity_mean <- candy$winpercent[as.logical(candy$fruity)]
t.test(chocolate_mean, fruity_mean)
```
The mean of chocolate candy is higher than that of fruity candies.

Q12. Is this difference statistically significant?

Yes, the difference is statistically different.

#3. Overall Candy Rankings

Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```
Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
library(dplyr)
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```
Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```
# Time to add some useful color
```{r}
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols)
```
- Q17. What is the worst ranked chocolate candy?
The worst ranked chocolate candy is Sixlets.

- Q18. What is the best ranked fruity candy?
Starburst is the best ranked fruity candy.


#4. Taking a look at the price percent
```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_col(fill = my_cols) +
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps =50)
```
Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
```{r}
candy %>%
  mutate(winper_by_priceper = winpercent/pricepercent) %>%
  arrange(desc(winper_by_priceper))
```
Tootsie Roll Midgies	is the highest ranked in terms of winpercent for the least money.

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
```{r}
head(candy[order(desc((candy$pricepercent))),])
```
Nik L Nip is not only the most expensive candy type in the dataset but also the least popular among the top5 most expensive.


Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), xend = 0, col = "gray40")) +
  geom_point()
```
# 5. Exploring the Correlation Structure

```{r}
library(corrplot)
cir <- cor(candy)
corrplot(cir)
```
Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Fruity and chocolate are anti-correlated.

Q23. Similarly, what two variables are most positively correlated?
(Chocolate and bar) and (winpercent and chocolate) variables are the most positively correlated variables.


# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col = my_cols, pch=16)
```
Using ggplot for pca data:
```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) +
  aes(x = PC1, y = PC2, 
      size = winpercent/100,
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)
p
```
```{r}
library(ggrepel) 

p + geom_text_repel(size = 3.3, max.overlaps = 50, col = my_cols) +
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
Using plotly package:
```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
The fruity, hard and pluribus variables are picked up strongly by PC1 in the positive direction.
