---
title: "Class 5: Data visualization with ggplot"
author: "Sanket Garg (A59026686)"
format: pdf
---

## Using GGPLOT

```{r}
plot(cars)
```


to use ggplot2, we first need to install it on our computers. To do this, we will use the function 'install.packages()'.

to use ggplot, I need to spell out at least 3 things:
- data (the stuff I want to plot as a data.frame)
- aesthetics (aes() values - how the data map to the plot)
- geoms (how I want things drawn)

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x = speed, y = dist) + 
  geom_point() +
  geom_smooth(method='lm')
```

```{r}
  ggplot(cars) + 
  aes(x = speed, y = dist) + 
  geom_point() + 
labs(title="Speed and Stopping Distances of cars", x="Speed (MPH)", y="Stopping          Distance(ft)",subtitle = "Your informative subtitle text here",caption="Dataset: 'cars'")+
  geom_smooth(method = 'lm', se=FALSE) + 
  theme_bw()
```
How many genes are upregulated?
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
ncol(genes)
table(genes$State)
round(table(genes$State)/nrow(genes)*100,2)
```

Making a scatter plot for the Genes dataset
```{r}
  ggplot(genes) + 
  aes(x=Condition1, y=Condition2) + 
  geom_point()
```
Changing the color of the datapoints based on their State:
```{r}
p <- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) + 
  geom_point()
p
```
Adding title and changing the x and y labels:
```{r}
p + 
  scale_color_manual(values = c("blue","gray", "red")) +
  labs(title = "Gene Expression Changes Upon Drug Treatment", x="Control (No Drug)", y="Drug Treatment")
```
Downloading the gapminder dataset:
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
library(dplyr)
gapminder_2007 <- gapminder %>% filter(year==2007)
```
Plotting the gapminder dataset:
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,y=lifeExp) +
  geom_point(alpha=0.4)
```
Controlling the aesthetics using the population and the continent aspects of the table:
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.4)
```
Plotting the scatter plot by using population aspect for color
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,y=lifeExp, color=pop) +
  geom_point(alpha=0.4)
```
Controlling the size of the data points:
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap,y=lifeExp, size=pop) +
  geom_point(alpha=0.4) +
  scale_size_area(max_size = 10)
```
Plotting for 1957
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)
ggplot(gapminder_1957)+
  aes(x=gdpPercap, y=lifeExp, color = continent, size = pop) +
  geom_point(alpha = 0.7) + 
  scale_size_area(max_size = 15)
```
Comparing 1957 and 2007:
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)
ggplot(gapminder_1957)+
  aes(x=gdpPercap, y=lifeExp, color = continent, size = pop) +
  geom_point(alpha = 0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

